### 2015-08-02 Harris_2015/manuscript-materials/draft.Rmd
x

\chead{\textit{Species interactions in Markov networks}}

@@

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
decades of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
another can become lost in the complex web of direct and indirect interactions
in real assemblages. For example, the competitive interaction between the
two shrub species in Figure 1A can become obscured by their shared tendency
to occur in unshaded areas (Figure 1B). If this sort of effect is common, then
the vast majority of ecologists' methods, which rely on test statistics
describing the total number of co-occurrences or overall correlation between
putatively interacting species [@diamond_island_1975; @strong_ecological_1984;
@gotelli_empirical_2009; @veech_probabilistic_2013; @pollock_understanding_2014]
will not generally provide much evidence regarding species' direct effects on
one another.

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumptions that all the pairwise interactions are zero and thus don’t need to
be controlled for.  Nevertheless, null models have dominated this field for more
than three decades [@strong_ecological_1984; @gotelli_empirical_2009].

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected
graphical models also known as Markov random fields;
@murphy_machine_2012] can provide a framework for understanding
the landscape-level consequences of pairwise species interactions, and
 for detecting them from observed presence-absence matrices. Markov
networks have been used in many scientific fields for decades in similar
contexts, from physics [where nearby particles interact magnetically
with one another; @cipra_introduction_1987] to spatial statistics [where
adjacent grid cells interact with one another; @harris_contact_1974;
@gelfand_modelling_2005]. While community ecologists explored
some related approaches in the 1980's [@whittam_species_1981],
computational limitations had previously imposed severe approximations
that produced unintelligible results [e.g. "probabilities" greater than one;
@gilpin_factors_1982].  Now that it is computationally feasible to fit these
models exactly, this kind of approach is worth a second look.

The rest of the paper proceeds as follows. First, I discuss how Markov
networks operate and how they can be used to simulate landscape-level
data or to predict the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the "true"
ecological structure is known, I compare this approach with several
existing methods for detecting species interactions. Finally, I discuss
opportunities for extending the approach presented here to larger
problems in community ecology.

@@

#### Markov networks.
Markov networks provide a method for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. In this section, I show how a set
of conditional relationships can be used to determine the kinds of species
assemblages that can occur. Methods for estimating conditional relationships
from presence-absence data are discussed in the next section.

Given a set of binary outcomes for each species (i.e. 0 for absence and 1 for
presence), a Markov network defines the relative probability of observing a
given presence-absence vector, $\vec{y}$, as
$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$

Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). Because the relative
probability of a presence-absence vector increases when positively-associated
species co-occur and decreases when negatively-associated species co-occur, the
model tends to produce assemblages that have many pairs of positively-associated
species and relatively few pairs of negatively-associated species (exactly as an
ecologist might expect).

Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then species'
marginal association rates can differ from this expectation. Predicting the
marginal relationships between species from their conditional interactions
entails summing over the different possible assemblages (Figure 2B), and the
number of possible assemblages can be intractably large, though some methods
beyond the scope of this paper can be employed to keep the calculations feasible
[@lee_learning_2012; @salakhutdinov_learning_2008]. Alternatively, as noted
below, some common linear and generalized linear methods can also be used as
computationally efficient approximations to the full network
[@lee_learning_2012; @loh_structure_2013].

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
(i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the
*rosalia* package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters.

@@

I also simulated a second set of landscapes using a stochastic community model
based on generalized Lotka-Volterra dynamics, as described in the Appendix. In
these simulations, each species pair was randomly assigned to either compete for
a portion of the available carrying capacity (negative interaction) or to act as
mutualists (positive interaction).  In these simulations, mutualism operated by
mitigating the effects of intraspecific competition on each partner's death
rate.  For these analyses, I simulated landscapes with up to 20 species and 25,
200, or 1600 sites (50 replicates per landscape size; see Appendix).

@@

For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
as described above. For the analyses with 20 species, I added a very weak
logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2 to
ensure that the model estimates were always finite. The bias introduced by this
prior should be small: the 95% credible interval on $\beta$ only requires that
one species' effect on the odds of observing a different species to be less than
a factor of 1500 (which is not much of a constraint). The logistic distribution
was chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but unlike
the Laplace distribution it is differentiable everywhere and does not force any
estimates to be exactly zero.

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method) to calibrate the performance of the Markov
network as a method for estimating species interactions. The first two
alternative interaction measures were the sample correlations and the partial
correlations between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
are undefined for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in the
corpcor package’s `pcor.shrink` function with the default settings
[@schafer_corpcor_2014]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products
[@friedman_sparse_2008].

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models [@gelman_weakly_2008] for each species,
using the other species on the landscape as predictors. To avoid the
identifiability problems associated with directed cyclic graphs
[@schmidt_modeling_2012], I then symmetrized the pairwise relationships within
species pairs via averaging.

The fourth alternative method, described in @gotelli_empirical_2009, involved
simulating possible landscapes from a null model that retains the row and column
sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
computed by the Pairs software described in @gotelli_empirical_2009 as my
null model-based estimator of species interactions.

The last two methods used the latent correlation matrix estimated by the
BayesComm package [@golding_bayescomm_2015] to evaluate the claim from recent
papers that the correlation coefficients estimated by "joint species
distribution models" provide an accurate assessment of species’ pairwise
interactions [@pollock_understanding_2014; see also @harris_generating_2015].  I
used both the average correlation between each pair of species’ latent
variables across 1000 Monte Carlo samples, as well as the average *partial*
correlation, as BayesComm's two estimators of species interactions.

@@

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths among all of the species that
were observed to occur across the simulated landscapes. This kind of evaluation
is important because ecologists need a consistent way to know how strong species
interactions are regardless of which species were present in a particular data
set or how many observations were taken. Because the different methods mostly
describe species interactions on different scales (e.g. correlations versus $Z$
scores versus regression coefficients), I used linear regression through the
origin to rescale the different estimates produced by each method so that they
had a consistent interpretation. After rescaling each method’s estimates, I
calculated squared errors between the scaled interaction estimates and “true”
interaction values across all the simulated data sets. These squared errors
determined the proportion of variance explained for different combinations of
method and landscape size (compared with a baseline model that assumed all
interaction strengths to be zero).

@@

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the correlation
between these species was positive in 94% of replicates, and the
randomization-based null model falsely reported positive associations 100% of
the time. Worse, more than 98% of these false conclusions were statistically
significant. The partial correlation and Markov network estimates, on the other
hand, each correctly identified the direct negative interaction between the
shrubs 94% of the time (although the confidence intervals overlapped zero in
most of the replicates).

#### Twenty species.
The accuracy of the evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3A). In general,
however, there was a consistent ordering, with the four methods that controlled
for indirect effects of species interactions performing the best: the Markov
network explained the largest portion of the variance (34.6% overall), followed
by the generalized linear models (29.9%), partial correlations from the raw
presence-absence data (27.5%), and partial correlations from BayesComm, the
joint species distribution model (26.1%).  The benefit of choosing the full
Markov network over the other three methods was largest on the smaller
landscapes, which are also the ones that are most representative of typical
analyses in this field [@gotelli_empirical_2009].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (18.6%), followed by the null model
(15.4%) and BayesComm's correlation matrix (11.6%). Although these last three
methods had different $R^2$ values, there was a close mapping between
their estimates (after controlling for the size of the simulated landscapes;
Figure 3B).  This suggests that the correlation matrices from joint species
distribution models and especially the effect sizes from null models only
contain noisy versions of the information that could be obtained more easily by
calculating correlation coefficients between species' presence-absence vectors.

@@

It’s important to note that none of these methods can identify the exact nature
of the pairwise interactions (e.g. which species in a positively-associated
pair is facilitating the other), particularly when real pairs of species can
reciprocally influence one another in multiple ways simultaneously
[@bruno_inclusion_2003]; with compositional data, there is only enough
information to provide a single number describing each species pair. To estimate
asymmetric interactions, such as commensalism or predation, ecologists would
need other kinds of data, as from time series, behavioral observations,
manipulative experiments, or natural history. These other sources of information
could also be used to augment the likelihood function with an informative prior
distribution, which could lead to better results on some real data sets than
was shown in Figure 3A.

Despite these limitations, Markov networks have enormous potential to improve
ecological understanding. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flashpoint for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated
by subtracting this value and the observed prevalence (cf Figure 2D).

Markov networks---particularly the Ising model for binary networks---have been
studied in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties, capabilities, and limits are well-understood in a
huge range of applications. Modeling species interactions using the same
framework would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
Markov networks can be fit with a mix of discrete and continuous variables,
and likely states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012]. There are even methods
[@whittam_species_1981; @tjelmeland_markov_1998] that would allow one species
to affect the sign or strength of the relationship between two other species,
tipping the balance between facilitation and exploitation
[@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
particular null model must be interpreted with care
[@roughgarden_competition_1983]. In complex ecological networks (and even in
small networks with three species), it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3B).
Estimating pairwise coefficients directly seems like a much more promising
approach: to the extent that the models' relative performance on real data sets
is similar to the range of results shown in Figure 3A, scientists in this field
could often double their explanatory power by switching from null models to
Markov networks (or increase it by nearly as much with linear or generalized
linear approximations). Regardless of the specific methods ecologists ultimately
choose, the most important consideration is clearly that we need to control for
indirect effects and estimate conditional relationships between species, rather
than marginal ones.

#### Acknowledgements:
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers also provided
useful feedback.

@@

**Figure 1.**
**A.** A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, as indicated by their negative coefficient (circled), but this effect
is substantially weaker. **B.** In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in locations
without trees makes their occurrence vectors positively correlated (circled).
**C.** Controlling for the tree species' presence with a conditional method
such as a partial covariance or a Markov network allows us to correctly
identify the negative interaction between these two species (circled).

**Figure 2.**
**A.** A small Markov network with two species.  The depicted abiotic
environment favors the occurrence of both species ($\alpha >0$), particularly
species 2 ($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking
these two species implies that they co-occur less than expected under
independence.  **B.** Relative probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. The exponent includes
$\alpha_1$ whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and $\beta$ only when both are present
($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four relative
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** Using the
probabilities, we can find the expected frequencies of all possible
co-occurrence patterns between the two species of interest. **D.** If $\beta$
equaled zero (e.g. if the species no longer competed for the same resources),
then the reduction in competition would allow each species to increase its
occurrence rate and the deficit of co-occurrences would be eliminated.

@@

\setlength{\parskip}{2pt}

>>>>

\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
\renewcommand{\headrulewidth}{0pt}

@@

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of direct and
indirect interactions in real assemblages. For example, the competitive
interaction between the two shrub species in Figure 1A can become obscured by
their shared tendency to occur in unshaded areas (Figure 1B). If this sort of
effect is common, then the vast majority of ecologists' methods, which rely on
test statistics describing the total number of co-occurrences or overall
correlation between putatively interacting species [@diamond_island_1975;
@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
@pollock_understanding_2014] will not generally provide much evidence regarding
species' direct effects on one another. The goal of this paper is to resolve
this long-standing problem in community ecology.

While competition doesn't reliably leave the expected pattern in community
structure (low co-occurrence rates at the landscape level), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumptions that all the pairwise interactions are zero and thus don’t need to
be controlled for.  Nevertheless, null models have dominated this field for more
than three decades [@strong_ecological_1984; @gotelli_empirical_2009].

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a
framework for understanding the landscape-level consequences of pairwise species
interactions, and  for detecting them from observed presence-absence matrices.
Markov networks have been used in many scientific fields for decades in similar
contexts, from physics [where nearby particles interact magnetically with one
another; @cipra_introduction_1987] to spatial statistics [where adjacent grid
cells interact with one another; @harris_contact_1974; @gelfand_modelling_2005].
While community ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed severe
approximations that led to unintelligible results [e.g. "probabilities" greater
than one; @gilpin_factors_1982].

The rest of the paper proceeds as follows. First, I discuss how Markov networks
operate and how they can be used to simulate landscape-level data or to make
exact predictions regarding the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the "true"
ecological structure is known, I compare this approach with several existing
methods for detecting species interactions. Finally, I discuss opportunities for
extending the approach presented here to other problems in community ecology,
such as disentangling the effects of biotic and abiotic interactions on species
composition.

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. In this section, I show how a set
of conditional relationships can be used to determine the how groups of species
can co-occur. Methods for estimating conditional relationships from
presence-absence data are discussed in the next section.

A Markov network defines the relative probability of observing a given vector of
species-level presences (1s) and absences (0s), $\vec{y}$, as

\centering

$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$

\raggedright
\setlength{\parindent}{1em}

Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
might have a value of $+2$ for two mutualists, indicating that the odds of
observing one species are $e^2$ times higher its partner is present than when
it is absent. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else equal---to
produce assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an ecologist
might expect).

Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then species'
marginal association rates can differ from this expectation. Determining the
marginal relationships between species from their conditional interactions
entails summing over the different possible assemblages (Figure 2B). This
becomes intractable when the number of possible assemblages is large, though
several methods beyond the scope of this paper can be employed to keep the
calculations feasible [@lee_learning_2012; @salakhutdinov_learning_2008].
Alternatively, as noted below, some common linear and generalized linear methods
can also be used as computationally efficient approximations to the full network
[@lee_learning_2012; @loh_structure_2013].

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average (i.e. it matches the sufficient statistics of the model distribution
to the sufficient statistics of the data). For the analyses in this paper, I
used the *rosalia* package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters.

@@

I also simulated a second set of landscapes using a stochastic community model
based on generalized Lotka-Volterra dynamics, as described in the Appendix. In
these simulations, each species pair was randomly assigned to either compete for
a portion of the available carrying capacity (negative interaction) or to act as
mutualists (positive interaction).  Here, mutualisms operate by mitigating the
effects of intraspecific competition on each partner's death rate.  For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or 1600
sites (50 replicates per landscape size; see Appendix).

@@

For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
as described above. For the analyses with 20 species, I added a very weak
logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2 to
ensure that the model estimates were always finite. The bias introduced by this
prior should be small: the 95% credible interval on $\beta$ only requires that
one species' effect on the odds of observing a different species to be less than
a factor of 1500 (which is not much of a constraint). The logistic distribution
was chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but unlike
the Laplace distribution it is differentiable everywhere and does not force any
estimates to be exactly zero. To confirm that this procedure produced reliable
estimates, I evaluated it on 50 bootstrap replicates (Appendix).

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method). The first two alternative interaction
measures were the sample correlations and the partial correlations between each
pair of species' data vectors on the landscape [@albrecht_spatial_2001;
@faisal_inferring_2010]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014]. In the
context of non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. This sort of model is very common for estimating
relationships among genes and gene products [@friedman_sparse_2008].

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models [@gelman_weakly_2008] for each species,
using the other species on the landscape as predictors. To avoid the
identifiability problems associated with directed cyclic graphs
[@schmidt_modeling_2012], I then symmetrized the relationships within species
pairs via averaging.

The fourth alternative method, described in @gotelli_empirical_2009, involved
simulating new landscapes from a null model that retains the row and column
sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
computed by the Pairs software described in @gotelli_empirical_2009 as my null
model-based estimator of species interactions.

The last two estimators used the latent correlation matrix estimated by the
BayesComm package [@golding_bayescomm_2015] in order to evaluate the untested
claim from recent papers that the correlation coefficients estimated by "joint
species distribution models" provide an accurate assessment of species’ pairwise
interactions [@pollock_understanding_2014; see also @harris_generating_2015].  
In addition to using the posterior mean of the correlation coefficient
[@pollock_understanding_2014], I also used the posterior mean of the *partial*
correlation, which might provide a better estimate of species' conditional
associations.

@@

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths among all of the species pairs
from the simulated landscapes. This kind of evaluation is important because
ecologists need a consistent way to know how strong species interactions are
regardless of which species were present in a particular data set or how many
observations were taken. Because the different methods mostly describe species
interactions on different scales (e.g. correlations versus $Z$ scores versus
regression coefficients), I used linear regression through the origin to rescale
the different estimates produced by each method so that they had a consistent
interpretation. After rescaling each method’s estimates, I calculated squared
errors between the scaled interaction estimates and “true” interaction values
across all the simulated data sets. These squared errors determined the
proportion of variance explained for different combinations of method and
landscape size (compared with a baseline model that assumed all interaction
strengths to be zero).

@@

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the correlation
between these species was positive in 94% of replicates, and the
randomization-based null model falsely reported positive associations 100% of
the time. Worse, more than 98% of false conclusions from the null model were
statistically significant. The partial correlation and Markov network estimates,
on the other hand, each correctly isolated the direct negative interaction
between the shrubs from their positive indirect interaction 94% of the time
(although the confidence intervals overlapped zero in most replicates).

#### Twenty species.
The accuracy of the evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3A). In general,
however, there was a consistent ordering, with the four methods that controlled
for indirect effects of species interactions performing the best: the Markov
network explained the largest portion of the variance (35% overall), followed
by the generalized linear models (30%), partial correlations from the raw
presence-absence data (28%), and partial correlations from BayesComm, the
joint species distribution model (26%).  The benefit of choosing the full
Markov network over the other three methods was largest on the smaller
landscapes, which are also the ones that are most representative of typical
analyses in this field [@gotelli_empirical_2009].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (19%), followed by the null model
(15%) and BayesComm's correlation matrix (11%). Although these last three
methods had different $R^2$ values, there was a close mapping among
their estimates (after controlling for the size of the simulated landscapes;
Figure 3B).  This suggests that the correlation matrices from joint species
distribution models and especially the effect sizes from null models only
contain noisy versions of the same information that could be obtained more
easily and interpretably by calculating correlation coefficients between
species' presence-absence vectors.

The bootstrapped results confirmed that the rosalia package's estimates of
species' conditional relationships were reasonably insensitive to sampling
variation (Appendix), and that the above ranking of the different methods was
robust (Appendix).

@@

It’s important to note that none of these methods can identify the exact nature
of the pairwise interactions [e.g. which species in a positively-associated
pair is facilitating the other; @schmidt_modeling_2012], particularly when real
pairs of species can reciprocally influence one another in multiple ways
simultaneously [@bruno_inclusion_2003]; with compositional data, there is only
enough information to provide a single number describing each species pair.
To estimate asymmetric interactions, such as commensalism or predation,
ecologists would need other kinds of data, as from time series, behavioral
observations, manipulative experiments, or natural history. These other sources
of information could also be used to augment the likelihood function with an
informative prior distribution, which could lead to better results on some real
data sets than was shown in Figure 3A.

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they are less vulnerable than some
of the most commonly-used methods to mistakenly identifying positive species
interactions between competing species. They also provide a simple answer to the
question of how competition should affect a species' overall prevalence, which
was a major flashpoint for the null model debates in the 1980’s
[@roughgarden_competition_1983; @strong_ecological_1984]. Equation 1 can be used
to calculate the expected prevalence of a species in the absence of biotic
influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's
effect on prevalence in a Markov network can then be calculated by subtracting
this value and the observed prevalence (cf Figure 2D). This kind of insight
would have been difficult to obtain without a generative model that makes
predictions about the consequences of species interactions; null models (which
presume *a priori* that interactions do not exist) have no way to make such
predictions.

Markov networks---particularly the Ising model for binary networks---have been
studied in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
Markov networks can be fit with a mix of discrete and continuous variables,
and likely states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward way to incorporate whole networks of biotic
interactions into species distribution models. Estimating each species'
conditional relationship with both biotic and abiotic factors in the same model
should lead to a better understanding of both kinds of ecological process. There
are even methods [@whittam_species_1981; @tjelmeland_markov_1998] that would
allow one species to affect the sign or strength of the relationship between
two other species, tipping the balance between facilitation and exploitation
[@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
particular null model must be interpreted with care
[@roughgarden_competition_1983]. Even in small networks with three species, it
may simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose null
[@gotelli_empirical_2009], especially when the test statistic is effectively
just a correlation coefficient (Figure 3B). Simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the range
of results shown in Figure 3A, scientists in this field could often double their
explanatory power by switching from null models to Markov networks (or increase
it by nearly as much with linear or generalized linear approximations).
Regardless of the specific methods ecologists ultimately choose, the most
important consideration is clearly that ecologists need to control for indirect
effects and estimate conditional relationships between species, rather than
marginal ones.

#### Acknowledgements:
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided
very helpful feedback on the text.

@@

**Figure 1.**
**A.** A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, but more weakly (circled coefficient). **B.** In spite of the
competitive interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors positively
correlated (circled). **C.** Controlling for the tree species' presence with a
conditional method such as a partial covariance or a Markov network leads to
correct identification of the negative shrub-shrub interaction (circled).

**Figure 2.**
**A.** A small Markov network with two species.  The abiotic environment favors
the occurrence of both species ($\alpha >0$), particularly species 2
($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking these two
species implies that they co-occur less than expected under independence.
**B.** Relative probabilities of all four possible presence-absence combinations
for Species 1 and Species 2. The exponent includes $\alpha_1$ whenever Species 1
is present ($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the
exponent includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
about 18.5.  **C.** Using the probabilities, we can find the expected
frequencies of all possible co-occurrence patterns between the two species of
interest. **D.** If $\beta$ equaled zero (e.g. if the species no longer competed
for the same resources), then the reduction in competition would allow each
species to increase its occurrence rate and the co-occurrence deficit would
be eliminated.

@@
